Exploration of white wine quality by Sai Venkat Kotha
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(memisc)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wines <- read.csv("wineQualityWhites.csv", header  = T)
wines[1] <- NULL
```

The dataset being explored in this analysis contains information about the 
quality ratings of several white wines. Each instance contains a quality rating 
for the wine between 0(very bad) and 10(very excellent) and also provides 
information about various chemical properties of the wine.

# Structure of the data

```{r echo=FALSE, Structure}
str(wines)
```

It can be seen that the dataset contains 4898 instances and there are 
11 variables which describe the chemical properties of the wine. The variable 
'quality' provides a rating for the wine.

```{r echo=FALSE, Summary}
summary(wines)
```

# Exploring individual variables

```{r echo=FALSE, Univariate_Plots-1}
qplot(data = wines, x = fixed.acidity, binwidth = 0.5, 
      xlab = "fixed.acidity (gm/litre)") +
  scale_x_continuous(limits = c(3,15), breaks = seq(3,15,0.5))
  
```

The distribution of the counts of 'fixed.acidity' variable looks normal. Most
of the wines have a fixed.acidity value between 6 and 7.5 and there are very 
few wines whose whose fixed acidity is less than 5 or more than 9.

```{r echo=FALSE, Univariate_Plots-2}
qplot(data = wines, x = volatile.acidity, binwidth = 0.1, 
      xlab = "volatile.acidity (gm/litre)") +
  scale_x_continuous(limits = c(0,1.3), breaks = seq(0,1.3,0.1))
```

From the above histogram it can be noticed that the value of 'volatile.acidity'
is a decimal usually between 0 and 1. Most of the wines have a volatile.acidty
value of 0.2 or 0.3 and there are very few wines whose volatile.acidity value is
more than 0.5.

```{r echo=FALSE, Univariate_Plots-3}
qplot(data = wines, x = citric.acid, binwidth = 0.1, 
      xlab = "citric.acid (gm/litre)") +
  scale_x_continuous(breaks = seq(0,1.7,0.1))
```

The amount of citric acid in about half of the wines is 0.3 and about 75% of
the wines have a citric acid amount less than 0.4. There are also a few outliers
where the amount of citiric acid is more than 1.

```{r echo=FALSE, message = FALSE, warning = FALSE, Univariate_Plots-4}
qplot(data = wines, x = residual.sugar, binwidth = 1, 
      xlab = "residual.sugar (gm/litre)") +
  scale_x_continuous(limits = c(0,20), breaks = seq(0,20,1))
```

The distribution of the 'residual.sugar' variable looks skewed with almost half
of the wines have a residual.sugar value between 0 and 5. About 75% of the wines
have a residual.sugar value less than 10 and there are a few outliers whose
residual.sugar value is more than 20.

```{r echo=FALSE, message = FALSE, warning = FALSE, Univariate_Plots-5}
qplot(data = wines, x = residual.sugar, xlab = "residual.sugar (gm/litre)") +
  scale_x_log10()
```

Transformed the skewed distribution of 'residual.sugar' using log10 to 
acheive a better distribution. The transformed distribution looks bimodal with
peaks at around 3 and 9.

```{r echo=FALSE, message = FALSE, warning = FALSE, Univariate_Plots-6}
qplot(data = wines, x = chlorides, binwidth = 0.02, 
      xlab = "chlorides (gm/litre)") +
  scale_x_continuous(breaks = seq(0, 0.4, 0.02))
```

The 'chlorides' variable represents the amount of salt in the wine. Most of the 
wines have a chlorides amount of 0.04 and there are very few wines whose
chlorides amount is more than 0.06.

```{r echo=FALSE, message = FALSE, warning = FALSE, Univariate_Plots-7}
qplot(data = wines, x = free.sulfur.dioxide, binwidth = 10, 
      xlab = "free.sulfur.dioxide (gm/mL)") +
  scale_x_continuous(limits = c(0,150), breaks = seq(0,150,10))
```

The above histogram shows that most of the wines have a free sulfur dioxide 
level between 20 and 40. About 75% of the wines have free sulfur dioxide
levels between 2 and 45. There are less number of wines whose free sulfur
dioxide level is more than 60.

```{r echo=FALSE, message = FALSE, warning = FALSE, Univariate_Plots-8}
qplot(data = wines, x = total.sulfur.dioxide, binwidth = 20, 
      xlab = "total.sulfur.dioxide (gm/mL)") +
  scale_x_continuous(limits = c(0,300), breaks = seq(0,300,20))
```

The amount of total sulfur dioxide in wines have a wider range than free sulfur
dioxide as it is the sum of free and bound sulfur dioxide. The distribution
looks normal with most of the wines having a total sulfur dioxide levels
between 100 and 160. The average amount of total sulfur dioxide in the wines 
is 138.4 gm/litre.

```{r echo=FALSE, Univariate_Plots-9}
qplot(data = wines, x = alcohol, binwidth = 0.5, 
      xlab = "alcohol (% by volume)") +
  scale_x_continuous(limits = c(7.5,15), breaks = seq(7.5,15,0.5))
```

The value of lowest alcohol content in a wine is 8 and the value of highest 
alcohol content is around 14. Most of the wines have alcohol content between 9
and 10.5. About 75% of the wines have an alcohol content value between 8 and
11.5.

```{r echo=FALSE, Univariate_Plots-10}
qplot(data = wines, x = quality, binwidth = 1) +
  scale_x_continuous(limits = c(0,10), breaks = seq(0,10,1))
```

Most of the wines received a quality rating from 5 to 7. There are very few
wines with a quality rating more than 7 and there are no wines which received a
rating of 10. It can also be seen that the lowest quality rating received is 3.

# Analysis of individual variables

### Structure of the dataset
The dataset contains 4898 instances of quality ratings received by various
white wines. Each instance also contains other variables which represent the 
chemical factors of the wine. These variables are 'fixed.acidity', 
'volatile.acidity', 'citric.acid', 'residual.sugar', 'chlorides', 
'free.sulfur.dioxide', 'total.sulfur.dioxide', 'density', 'pH', 'sulphates', 
'alcohol'. All the variables other than quality are represented by 
decimal values. 

About 50% of the wines received a rating of 6 or lower and only 25% of wines
received a rating greater than 6. Most of the wines have alcohol content 
between 9 and 11.

### Main features of interest in the dataset
The main features of interest in the dataset are alcohol and 
quality. I would like to dtermine if the amount of alcohol affected the 
rating of a wine.

### Other features in the dataset that might help support the \
investigation into the features of interest
The other features that could support the investigation are residual.sugar, 
density and total sulfur dioxide. The amount of residual sugar in wine that 
sweetens the wine could have affected the rating of the wine. The amount of 
total sulfur dioxide is also important as it pevents the oxidation of the wine. 
Excess oxidation causes the wine to degrade and lose its aroma and taste.

### Unusual distributions and data transformations
The distribution of the amount of sugar represented by residual.sugar variable
had a skewed distribution. Transforming this distribution using log10 
transformation changed the distribution to bimodal with peaks at 3 and 9.

# Analyzing relationships among two variables

### Correlation coefficients

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots-1}
cor(wines)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots-2}
ggplot(data = wines, aes(x = factor(quality), y = alcohol)) +
  geom_jitter(alpha = 1/4) +
  geom_boxplot(col = "green") + 
  stat_summary(fun.y = mean, geom = "point", shape = 4, color = "red") +
  labs(y = "alcohol (% by volume)", x = "quality")
```

The above plot indicates that the wines with higher quality have high amounts
of alcohol. The wines which received a lower quality rating from 3 to 5 have 
average alcohol amount of around 10% whereas the wines with higher quality 
ratings have average alcohol amount more than 11%.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots-3}
ggplot(data = wines, aes(x = factor(quality), y = density)) +
  geom_jitter(alpha = 1/4) +
  geom_boxplot(col = "green") +
  scale_y_continuous(limits = c(0.985,1.005)) +
  stat_summary(fun.y = mean, geom = "point", shape = 4, col = "red") +
  labs(y = "density (gm/mL)", x = "quality")
```

It can be noticed that the wines with higher quality have lower density than
that of the wines with lower quality. The density of almost all the wines is 
within the range 0.985 to 1.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots-4}
ggplot(data = wines, aes(x = factor(quality), y = residual.sugar)) +
  geom_jitter(alpha = 1/4) +
  geom_boxplot(col = "green") +
  scale_y_continuous(limits = c(0,25), breaks = seq(0,25,5)) +
  stat_summary(fun.y = mean, geom = "point", col = "red", shape = 4) +
  labs(y = "residual.sugar (gm/litre)", x = "quality")
```

The amount of residual sugar appears to be in same amounts across all the 
qualities of the wine.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots-5}
ggplot(data = wines, aes(x = factor(quality), y = total.sulfur.dioxide)) +
  geom_jitter(alpha = 1/5) +
  geom_boxplot(col = "green") +
  scale_y_continuous(limits = c(0,300)) +
  stat_summary(fun.y = mean, geom = "point", col = "red", shape = 4) +
  labs(y = "total.sulfur.dioxide (mg/litre)", x = "quality")
```

The amount of total sulfur dioxide does not seem to have any relationship with 
the quality of the wine. The wines of all quality have total sulfur dioxide 
amounts in the same range between 50 and 250.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots-13}
ggplot(data = wines, aes(x = residual.sugar, y = total.sulfur.dioxide)) +
  geom_point(alpha = 1/5) +
  labs(x = "residual.sugar (gm/litre)",y = "total.sulfur.dioxide (mg/litre)") +
  scale_x_continuous(limits = c(0,30), breaks = seq(0,30,5)) +
  scale_y_continuous(limits = c(0,300), breaks = seq(0,300,50)) +
  geom_smooth(method = lm, se = FALSE, color = 'red')
```

The relationship is not very strong between residual sugar and total sulfur 
dioxide. There are a large number of wines with residual sugar amount less than
5 and having a total sulfur dioxide amount between 50 and 200. As the amount of
residual sugar in wine increases there is a slight increase in the amount of
total sulfur dioxide as well.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots-14}
ggplot(data = wines, aes(x = residual.sugar, y = density)) +
  geom_point(alpha = 1/10) + 
  labs (x = "residual.sugar (gm/litre)", y = "density (gm/mL)") +
  scale_x_continuous(limits = c(0,30), breaks = seq(0,30,5)) +
  scale_y_continuous(limits = c(0.985,1.005)) +
  geom_smooth(method = lm, se = FALSE, color = 'red')
```

The relationship between residual.sugar and density looks almost linear. The
wines that have more amounts of residual sugar also have a higher density than 
the wines with lower amounts of residual sugar.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots-15}
ggplot(data = wines, aes(x = residual.sugar, y = alcohol)) +
  geom_point(alpha = 1/3) +
  labs (x = "residual.sugar (gm/litre)", y = "alcohol (% by volume)") +
  scale_x_continuous(limits = c(0,30), breaks = seq(0,30,5)) +
  geom_smooth(method = lm, se = FALSE, color = 'red')
```

There is no strong relationship between residual.sugar and alcohol. But it can
be noticed that as the amount of alcohol in the wine increased the variance in
the amount of residual sugar decreased. And the wines with highest alcohol 
amount had residual sugar amount of around 10 gm/litre whereas there are wines 
with lower alcohol amount with residual sugar amount of 20 gm/litre.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots-16}
ggplot(data = wines, aes(x = alcohol, y = density)) +
  geom_point(alpha = 1/10) +
  labs (x = "alcohol (% by volume)", y = "density (gm/mL)") +
  scale_y_continuous(limits = c(0.985,1.005)) +
  geom_smooth(method = lm, col = 'red')
```

The above plot reveals a slight negative linear relationship between alcohol 
and density. It can be seen that the wines with high amounts of alcohol have
less density that that of the wines with low amounts of alcohol.

# Analysis of bi-variate relationships

### Relationships observed that involved features of interest
The quality rating received by a wine did not have a strong relationship with
any other variables. Quality of a wine was compared with variables like alcohol,
density, residual.sugar and total.sulfur.dioxide. Almost all the wines had the 
same range of values for these other variables.

To find how the mean and median values of these variables compared to the
quality of wine, I have created a new variable named "quality_class". The wines 
with quality rating less than or equal to 4 belong to the class "low", the wines
with quality rating 5 or 6 belong to the class "medium" and the wines with 
quality rating higher than 6 belong to the class "high".

It was observed that the wines with higher quality had more amounts of alcohol
and less density than the wines with lower quality. The amounts of residual
sugar was almost all the same in all the quality classes of wine.

### Relationships observed among other features
The other features explored were residual.sugar, density and alcohol. There is
no strong relationship between residual.sugar and alcohol but density has 
strong relationships with both residual sugar and alcohol. There is a positive
linear relationship between residual sugar and density and there is a negative
linear relationship between alcohol and density.

### Strongest relationships observed
The strongest relationship found was between residual sugar and density and 
also between alcohol and density.

# Relationships among multiple variables

```{r echo=FALSE, message=FALSE, warning=FALSE}
wines$quality_class <- NA
wines <- transform(wines, quality_class = ifelse(quality<=4, "low",
                                          ifelse(quality==5 | quality ==6,
                                                 "medium","high")))
wines$quality_class <- factor(wines$quality_class, levels = 
                                c("low","medium","high"))
```


```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots-1}
ggplot(data = wines, aes(x = residual.sugar, y=density, col=factor(quality))) +
  geom_point(alpha = 1/10) +
  scale_x_continuous(limits = c(0,25), breaks = seq(0,25,5)) +
  scale_y_continuous(limits = c(0.985,1.005)) +
  geom_smooth(method = "lm", se = FALSE, size=1) +
  scale_color_brewer(type='seq', guide=guide_legend(title='Quality')) +
  labs(x = "residual.sugar (gm/litre)", y = "density (gm/mL)")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = wines, aes(x = round(residual.sugar), y = density)) +
  geom_line(aes(col = quality_class), stat = "summary", fun.y = mean) +
  scale_x_continuous(limits = c(0,30), breaks = seq(0,30,5)) +
  scale_y_continuous(limits = c(0.985,1.005)) +
  #scale_color_brewer(type='seq', guide=guide_legend(title='Quality Class')) +
  labs(x = "residual.sugar (gm/litre)", y = "density (gm/mL)")
```

The wines that received a higher quality rating appear to  have lower density at 
lower amounts of residual sugar when compared to the wines with lower quality 
ratings.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots-2}
ggplot(data = wines, aes(x = alcohol, y =density, col=factor(quality))) +
  geom_point(alpha = 1/10) +
  scale_y_continuous(limits = c(0.985,1.01)) +
  geom_smooth(method = "lm", se = FALSE, size=1) +
  scale_color_brewer(type='seq', guide=guide_legend(title='Quality')) +
  labs(x = "alcohol (% by volume)", y = "density (gm/mL)")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = wines, aes(x = round(alcohol), y = density)) +
  geom_line(aes(col = quality_class), stat = "summary", fun.y = mean) +
  scale_y_continuous(limits = c(0.985,1.01)) +
  labs(x = "alcohol (% by volume)", y = "density (gm/mL)")
```

As we have already noticed that wines with higher quality ratings have lower 
density and high amounts of alcohol, this fact is again revealed by this plot. 
It can be noticed that until 10% of alcohol by vloume, the wines with high 
quality have higher density but after that these wines have the lowest density 
than the other wines.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots-3}
ggplot(data = wines, aes(x = residual.sugar, y=alcohol, col=factor(quality))) +
  geom_point(alpha = 1/5) +
  scale_x_continuous(limits = c(0,30), breaks = seq(0,30,5)) +
  geom_smooth(method = "lm", se = FALSE, size=1) +
  scale_color_brewer(type='seq', guide=guide_legend(title='Quality')) +
  labs(x = "residual.sugar (gm/litre)", y = "alcohol (% by vokume)")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = wines, aes(x = round(residual.sugar), y=alcohol)) +
  geom_line(aes(col = quality_class), stat = "summary", fun.y = mean) +
  scale_x_continuous(limits = c(0,30), breaks = seq(0,30,5)) +
  labs(x = "residual.sugar (gm/litre)", y = "alcohol (% by vokume)")
```

As the previous plots indicated, the wines with higher quality ratings
have higher amounts of alcohol at lower amounts of residual sugar. It can also 
be noticed that the amount of alcohol in the wines belonging to medium quality
ratings suddenly increase at higher amounts of residual sugar.

# Multivariate Analysis

### Relationships observed
This part of the analysis did not reveal any new features or patterns, instead
it strengthened the patterns found in the previous parts of the analysis. As we
have already seen that wines with high quality have higher amounts of alcohol
and lower density, surprisingly it can be seen from the plot density vs alcohol
that wines with high quality have higher density than wines with lower quality
until 10% alcohol by volume. After 10% alcohol by volume, the density of the
wines is less than the lower quality wines.

------

# Final Plots and Summary

### Plot One

```{r echo=FALSE, Plot_One}
theme_update(plot.title = element_text(hjust = 0.5))
ggplot(data = wines, aes(x = factor(quality), y = alcohol)) +
  geom_jitter(alpha = 1/4) +
  geom_boxplot(col = "green") + 
  stat_summary(fun.y = mean, geom = "point", shape = 4, color = "red") +
  labs(y = "alcohol (% by volume)", x = "quality") +
  ggtitle("alcohol vs quality")
```

### Description One
This plot indicates that wines with higher quality have more amounts of
alcohol than in the wines with lower quality. Even though the difference in 
the means of the amount of alcohol is not very big, it is still considerable.

### Plot Two

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ggplot(data = wines, aes(x = residual.sugar, y = density)) +
  geom_point(alpha = 1/10) + 
  labs (x = "residual.sugar (gm/litre)", y = "density (gm/mL)") +
  scale_x_continuous(limits = c(0,30), breaks = seq(0,30,5)) +
  scale_y_continuous(limits = c(0.985,1.005)) +
  geom_smooth(method = lm, se = FALSE, color = 'red') +
  ggtitle("density vs residual sugar")
```

### Description Two
This plot shows a linear relationship between the amount of residual sugar and
the density of the wine. The reason for this linear relationship could be the 
fact that density of wine is measured based on the percentage of alcohol and 
the sugar content of the wine.

### Plot Three

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
ggplot(data = wines, aes(x = alcohol, y =density, col=factor(quality))) +
  geom_point(alpha = 1/10) +
  scale_y_continuous(limits = c(0.985,1.01)) +
  geom_smooth(method = "lm", se = FALSE, size=1) +
  scale_color_brewer(type='seq', guide=guide_legend(title='Quality')) +
  labs(x = "alcohol (% by volume)", y = "density (gm/mL)") +
  ggtitle("density vs alcohol and quality")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = wines, aes(x = round(alcohol), y = density)) +
  geom_line(aes(col = quality_class), stat = "summary", fun.y = mean) +
  scale_y_continuous(limits = c(0.985,1.01)) +
  labs(x = "alcohol (% by volume)", y = "density (gm/mL)") +
  ggtitle("density vs alcohol and quality class")
```

### Description Three
Thee above two plots look interesting because the findings prior to this plot 
indicated that usually the wines that belonged to higher quality class have 
slightly high amount of alcohol and lower density. But this plot shows that for 
a particular alcohol content, the density of the wines with higher quality
rating is higher than the wines with lower quality rating. But this trend is 
only observed at alcohol amounts between 8 and 10. At alcohol amounts greater 
than 10, the wines with higher quality rating have lower density.

------

# Reflection
The white wine dataset contains 4898 instances of quality rating received by
various wines. Each instance also contains 11 other variables based on the 
physicochemical tests. I started the EDA of this dataset by looking at the
distribution of these individual variables. All the variables except the quality
variable had numerical values. The distribution of residual.sugar variable was
skewed and I transformed the variable using log transformation.

The variables did not have strong correlation coefficients and the only 
variables pairs which had some reasonable correlation were residual.sugar & 
alcohol, residual.sugar & density, density & alcohol, quality & alcohol. 
Exploring these pairs did not result in any strong relationships apart from 
residual.sugar vs density and density vs alcohol which had linear relationships.
This lack of correlation among variables caused trouble as the plots did not
reveal any proper relationships to explore further.

To determine how wines with similar quality rating compared to each other, I 
created a new variable called "quality_class" with values "low", "medium" and 
"high". Exploring these classes indicated that the wines that belonged to the
'high' quality class had lower density and high amounts of alcohol when 
compared to the other lower classes of wines. The amount of residual sugar
was almost the same in the wines of all the classes. Dividing the wines into
these classes and exploring them turned out to be a success as it helped in 
observing the trends in alcohol and density among the wines. 

The exploratory analysis performed so far did not result in any strong 
relationships that could be used for building a predictive model. The reason
for this could be the lack of important characteristics of wine in the dataset.
The analysis could be improved if the dataset also had other characteristics
such as the type of soil in which the grapes used were grown, the species of
grapes used, the climate in which the wine was made etc.